---
title: "US coal exports"
author: "Campbell Miller and Aidan Bergsman"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: github_document
fig_width: 10
fig_height: 10
---

```{r setup, include=FALSE}
## These next lines set the default behaviour for all R chunks in the .Rmd document.
## I recomend you take a look here: https://rmarkdown.rstudio.com/authoring_rcodechunks.html
knitr::opts_chunk$set(
  echo = TRUE, ## Show all R output
  cache = TRUE, ## Cache the results to increase performance.
  message = FALSE ## Suppress messages (e.g. about namespace conflicts)
  ) 

```

## Preliminaries: 

### Load libraries

It's a good idea to load your libraries at the top of the Rmd document so that everyone can see what you're using. Similarly, it's good practice to set `cache=FALSE` to ensure that the libraries are dynamically loaded each time you knit the document.

*Hint: I've only added the libraries needed to download and read the data. You'll need to load additional libraries to complete this assignment. Add them here once you discover that you need them.* 

```{r libs, cache=FALSE}
## Install the pacman package if necessary
if (!require("pacman")) install.packages("pacman")
## Install other packages using pacman::p_load()
pacman::p_load(httr, readxl, here, data.table, tidyverse, janitor, ggplot2, lubridate, zoo, forcats,ggthemes,viridis)
```

### Read in the data

Use `httr::GET()` to fetch the EIA excel file for us from web. (We'll learn more about `httr`, GET and other HTTP methods when we get to webscraping next week.) 

```{r get}
# library(here) ## Already loaded
# library(httr) ## Already loaded
url = "https://www.eia.gov/coal/archive/coal_historical_exports.xlsx"
## Only download the file if we need to
if(!file.exists(here::here("data/coal.xlsx"))) {
  GET(url, write_disk(here::here("data/coal.xlsx")))
}
```

Next, we read in the file.

```{r read}
# library(readxl) Already loaded
coal = read_excel(here::here("data/coal.xlsx"), skip = 3, na = ".")
```

We are now ready to go.

## 1) Clean the column names

The column (i.e. variable) names aren't great: Spacing, uppercase letters, etc. 
```{r names, dependson=coal}
  names(coal)
##Use clean_names to put them in lower case and _ instead of space format
coal = coal %>%
clean_names()
```

Clean them. 

*Hint: Use either `gsub()` and regular expressions or, more simply, the `janitor()` package. You will need to install the latter first.*


## 2) Total US coal exports over time (year only)

Plot the US's total coal exports over time by year ONLY. What secular trends do you notice in the data?

*Hints: If you want nicely formatted y-axis label, add `+ scale_y_continuous(labels = scales::comma)` to your `ggplot2` code.*

**Please put your (verbal) answers in bold.**

```{r}
##create graph showing US total coal exports over year
ggplot(coal, aes(x = year, y = total)) +
  geom_bar(stat = "identity", 
           fill = 'chocolate4') +
   scale_y_continuous(labels = scales::comma) +
 theme(panel.background = element_rect(fill = 'cornsilk3', 
                                         colour = 'black')) +
  xlab("Year") + 
   ylab("Total Coal Exports from the US") + 
  ggtitle("Total Coal Exports from the US by Year") +
    theme(plot.title = element_text(hjust = 0.4)) 

```


**A secular trend is a variable that shows a consistent pattern over a given period of time. The trend shown in this graph seems to be a growth in total coal exports from the US. Before 2007 there is a consistent growth in coal exports every year and after that there is more variation with significant increases and decreases in exports. The exports oscillate around 80,000,000 coal beginning around 2007 in which  the level cycles around this value. The oscillations do appear to occur regularly after 2007 which could show new coal deposits being found which results in massive exports until supplies run low and the total exports dip back down until a new deposit is found and the total value exported massively increases again.**

## 3) Total US coal exports over time (year AND quarter)

Now do the same as the above, except aggregated quarter of year (2001Q1, 2002Q2, etc.). Do you notice any seasonality that was masked from the yearly averages?

*Hint: ggplot2 is going to want you to convert your quarterly data into actual date format before it plots nicely. (i.e. Don't leave it as a string.)*

```{r,fig.width= 11}
#create a data.table for the data
coal_datatable = as.data.table(coal)

# unite year and quarter variables into "date" column
coal_datatable = coal_datatable %>% unite(date, c("year", "quarter"), sep = "-")

coal_datatable[,q_total := sum(total,na.rm = TRUE),by = date]

coal_datatable[,date_2 := as.yearqtr(date),]

# make graph

ggplot(coal_datatable, aes(x = date_2, y = q_total)) +
 geom_bar(stat = "identity", fill = 'chocolate4') +
  scale_y_continuous(labels = scales::comma) +
  theme(panel.background = element_rect(fill = 'cornsilk3', 
                                         colour = 'black')) +
  xlab("Year and Aggregated Quarter") + 
   ylab("Coal Exports from the US") + 
   ggtitle("Total Coal Exports from the US Aggregated by Quarter") +
  theme(plot.title = element_text(hjust = 0.4)) 

```

**There is some seasonality that was masked from the yearly averages. In the previous graph the initial years through 2007 were only increasing in total coal exports, but in the new graph we can see some variety in these years as values increase and decrease throughout the quarters. Likewise in the initial graph from year 2012 through 2016 the values were strictly decreasing, but in the new graph we can see there are quarters that increase from the previous quarter. The general trends from the initial graph remain but the quarter aggregated graph could allow us to make some conclusions as to how different quarters affect annual coal exports. For example it seems that a majority of the first quarter portions of the graph are a lower value than the previous fourth quarter and the following second quarter, which is typically higher. This trend could be looked upon with information about coal production/mining to see if this conclusion makes sense with what time of year coal is most heavily produced.**

## 4) Exports by destination country

### 4.1) Create a new data frame

Create a new data frame called `coal_country` that aggregates total exports by destination country (and quarter of year). Make sure you print the resulting data frame so that it appears in the knitted R markdown document.
```{r}
#create datatable where you sum total coal by the destination country and the new quarter and year date.
coal_country = coal_datatable [, list(total_coal = sum(total, na.rm = TRUE)), by = list(coal_destination_country, date_2)]

#order the new datatable by destination country and quarter and print with []
setorder(coal_country, coal_destination_country, date_2)[]
```


### 4.2) Inspect the data frame

It looks like some countries are missing data for a number of years and periods (e.g. Albania). Confirm that this is the case. What do you think is happening here?

```{r}
# showing count of quarterly data observations for each country
count(coal_country, coal_destination_country)[]

# Showing count of countries with less than half (40) of quarterly data observations
count(coal_country,coal_destination_country) %>% filter(n<40)
```

**The United Kingdom has 80 occurrences of quarterly data in the data set which is the maximum that can occur (20 years times 4 quarters equals 80). There are many countries that do not have this maximum of 80, such as Albania with one observation or Vietnam with 39. In fact, 94 countries (which is roughly 62% of the observations) have less than half of the 80 total observations they can have over this time period. What this likely means is that the US would not record an observation if they did not export to a certain country for a given quarter rather than tracking it as a 0 in exports. This would mean that observations are only in the data set if the US actually exported to them in that quarter. Because of this, there are many "na" values in the data set for when the US did not export coal to certain countries.**

### 4.3) Complete the data frame

Fill in the implicit missing values, so that each country has a representative row for every year-quarter time period. In other words, you should modify the data frame so that every destination country has row entries for all possible year-quarter combinations (from 2002Q1 through the most recent quarter). Order your updated data frame by country, year and, quarter. 

*Hints: See `?tidyr::complete()` for some convenience options. Again, don't forget to print `coal_country` after you've updated the data frame so that I can see the results.*

```{r}
# expand coal_country to include all combinations of country and year/quarter
coal_country_expanded = 
  complete(coal_country, coal_destination_country, date_2, 
           fill = list(total_coal = 0))

# check to see that it worked (should have n=80 for all countries)
count(coal_country_expanded, coal_destination_country)[]

# print newly expanded coal_country data frame
coal_country_expanded
```



### 4.4 Some more tidying up

In answering the previous question, you _may_ encounter a situation where the data frame contains a quarter --- probably `r paste0(format(Sys.Date(), "%Y"), quarters(Sys.Date()-months(3)))` --- that is missing total export numbers for *all* countries. Did this happen to you? Filter out the completely missing quarter if so. Also: Why do you think this might have happened? (Please answer the latter question even if it didn't happen to you.) 

(Filter for the most recent quarter to see if we are missing total export numbers for all countries)

```{r}
# Filter data for most recent quarter (2021 Q 4)
coal_country_expanded %>% #group_by(date_2) %>%
  filter(date_2 == "2021 Q4") %>% filter(total_coal!=0)
```

**The latest date in our data set from the previous question is 2021 Q4. We filtered the data frame to include only observations from 2021 Q4 that had positive total coal values, which there are 54 of. As a result, we are not missing total coal numbers from all countries in the most recent quarter of the data set. The reason the situation might have happened is if even one country had an NA or 0 value for 2021 Q4 total coal imported, the way we coded the previous questions would have propagated a value for that date for every country with a total coal value of 0 for every country. This would have resulted in useless values in the date_2 column as the amount of coal for every country would be 0. If this problem occurred, We could fix it by filtering the data frame so it did not include 2021 Q4.**


### 4.5) Culmulative top 10 US coal export destinations

Produce a vector --- call it `coal10_culm` --- of the top 10 top coal destinations over the full `r min(coal[, which(grepl('Year|year', names(coal)))], na.rm=T)`--`r `max(coal[, which(grepl('Year|year', names(coal)))], na.rm=T)` study period. What are they?

```{r}
#create vector grouping by import country and summing the total coal import to see which imported the most
coal10_culm = coal_country_expanded %>%
  group_by(coal_destination_country) %>%
  summarize(top_coal_importers = sum(total_coal)) %>%
  arrange(desc(top_coal_importers))
coal10_culm

```

**The top 10 US coal export destinations are Canada, Netherlands, Brazil, India, South Korea, Japan, UK, Italy, China, and Germany.**


### 4.6) Recent top 10 US coal export destinations

Now do the same, except for most recent period on record (i.e. final quarter in the dataset). Call this vector `coal10_recent` and make sure to print it so that I can see it too. Are there any interesting differences between the two vectors? Apart from any secular trends, what else might explain these differences?

```{r}
#do the same vector as before but filter to just the most recent quarter
coal10_recent = coal_country_expanded %>%
  filter(date_2 == "2021 Q4") %>%
  group_by(coal_destination_country) %>%
  summarize(top_coal_importers = sum(total_coal)) %>%
  arrange(desc(top_coal_importers))
coal10_recent
```

**In comparing the two vectors, we can see that many countries overlap between them. For example, Canada, the Netherlands, Brazil, India, South Korea, Japan, China, and Germany are included in both top ten lists. What is interesting to note is that several countries in Asia rose to the top of the list for the most recent quarter, including China, India, and Japan comprising the top 3 nations the US exported coal to in 2021's 4th quarter. Something that may be able to explain these differences would be the amount of coal these countries have access to in their borders and how these levels have changed over the decades. It would make sense that countries would start to deplete their sources of coal over time, and as a result would need to import from other countries to maintain their levels. Because of this, countries that were lower on the import list started to climb higher in the most recent quarter. Also, countries like Morocco and Ukraine entered the top 10 and it is interesting to consider what caused this change. Recent research shows that Morocco's energy use comes from 90% coal, so it makes sense why they would be importing more coal in recent years. We can also see in the data that Ukraine had 0 imported coal for many of the early years of this data set, so it makes sense that even if it has high current imports it would not show up in the previous vector but could show up here.**


### 4.7) US coal exports over time by country

Plot the quarterly coal exports over time, but now disaggregated by country. In particular, highlight the top 10 (cumulative) export destinations and then sum the remaining countries into a combined "Other" category. (In other words, your figure should contain the time series of eleven different countries/categories.)

```{r, fig.width= 11}
#make list for the top ten countires from previous q
top10 = c("Canada", "Netherlands", "Brazil", "India", "South Korea (Republic of Korea)", "Japan", "United Kingdom", "Italy", "China", "Germany, Federal Republic of")

#make a new variable import_name where all countries other than top 10 are other and the top 10 retain their names. 
coal_new = coal_country_expanded %>%
  mutate(import_name = ifelse(coal_destination_country %in% top10, coal_destination_country, "Other"))

#plot based on name in other category
ggplot(coal_new, aes(x = date_2, y = total_coal, color = import_name)) +
  geom_line() +
  scale_y_continuous(labels = scales::comma) +
  xlab("Year and Aggregated Quarter") +
  ylab("Total Coal Imports from the US") +
  ggtitle("Coal Exports from the US to the Top 10 Importing Countries compared to all Others") +
   theme(plot.title = element_text(hjust = 0.35)) +
  labs(color = "Country Importing Coal") +
  scale_color_brewer(palette = "Paired") 

```


### 4.8) Make it pretty

Take your previous plot and add some swag to it. That is, try to make it as visually appealing as possible without overloading it with chart junk.

*Hint: You've got loads of options here. If you haven't already done so, consider a more bespoke theme with the `ggthemes`, `hrbrthemes`, or `cowplot` packages. Try out `scale_fill_brewer()` and `scale_colour_brewer()` for a range of interesting colour palettes. Try some transparency effects with `alpha`. Give your axis labels more refined names with the `labs()` layer in ggplot2. While you're at it, you might want to scale (i.e. normalise) your y-variable to get rid of all those zeros. You can shorten any country names to their ISO abbreviation; see `?countrycode::countrycode`. More substantively --- but more complicated --- you might want to re-order your legend (and the plot itself) according to the relative importance of the destination countries. See `?forcats::fct_reorder` or forcats::fct_relevel`.*

```{r, fig.width= 11}
# use forcats to reorder importing country name by total coal imported (to sort from most to least important)
test = coal_new %>% mutate(import_name = fct_reorder(import_name,total_coal, .desc = TRUE)) 

ggplot(test, aes(x = date_2, y = total_coal, color = import_name)) +
  geom_line(position = "stack") +
  scale_y_continuous(labels = scales::comma) +
  xlab("Year and Aggregated Quarter") +
  ylab("Total Coal Imports from the US") +
  ggtitle("Coal Exports from the US to the Top 10 Importing Countries and all Others") +
   theme(plot.title = element_text(hjust = 0.35),
         panel.background = element_rect(fill = "gray"),
         legend.title = element_text(size = 12))   + 
  scale_color_brewer(palette = "RdYlBu", name = "Country Importing") +
  geom_area(aes(fill = import_name), alpha = 0.8) +
  scale_fill_brewer(palette = "RdYlBu", name = "Country Importing")
```


## 5) Show me something interesting

There's a lot still to explore with this data set. Your final task is to show me something interesting. Drill down into the data and explain what's driving the secular trends that we have observed above. Or highlight interesting seasonality within a particular country. Or go back to the original `coal` data frame and look at exports by customs district, or by coal type. Do we changes or trends there? Et cetera. Et cetera. My only requirement is that you show your work and tell me what you have found.

```{r}
# graph china's quarterly change in coal imports to see if we can discover how it became the number 1 importer in most recent quarter.
coal_country_expanded %>% 
  filter(coal_destination_country == c("China")) %>% 
  ggplot(aes(x=date_2,y=total_coal)) + 
  geom_bar(stat = "identity",fill = "gray", color = "black") +
     scale_y_continuous(labels = scales::comma) +
  ylab("Total Coal Import") + 
  xlab("Date (Quarterly)")  +
  ggtitle("China's Quarterly Coal Imports From US") +
  theme_stata() +
  theme(plot.title = element_text(size=20)) 
```


**In a previous question, we discovered that China had imported the most coal from the US in the most recent quarter data was available for (2021 Q4). We wanted to dig deeper into the historical importing patterns in China to see if we could explain why they jumped into the top spot. According to the above graph, China has had large fluctuations in their coal imports from 2002 to 2021. Up until 2008, China was importing very little to no coal. From 2008 until around 2012, China was importing more coal, peaking at just under 6,000,000 units. Interestingly, this is followed by a sharp decline, where imports fluctuate below 1,000,000 units until China again increased demand at the end of 2020 and into 2021. Because of the multiple fluctuations in import numbers, we would think that the amount of coal the US exports to China is most dependent on political relations at any given time, rather than our earlier conjecture that countries import more coal when they begin to run low on domestic stocks.** 
